---
title: "Statistical inference with the GSS data"
output: 
  html_document: 
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

## Setup

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
library(statsr)
```

### Load data

Make sure your data and R Markdown files are in the same directory. When loaded
your data file will be called `gss`. Delete this note when before you submit 
your work. 

```{r load-data}
load("gss.Rdata")
```



* * *

## Part 1: Data

The data from the General Social Survey (GSS) consists of a dataframe with 114 variables and 57061 observations. Therefore, the sample size of this dataset would be 57061 with the population . Each variable is a societal opinion of individuals ranging over a variety of topics such as financial position, religious beliefs, abortion, race relations, crime, quality of life, drug use to name a few. 

The population of the data spans from 1972 to 2012 providing 4 decades worth of data covering a wide age group of 18-89. 

Generalizability: Since the data is collected randomly from households across the country random sampling is taking place and hence, the sample is generalizable to the entire US population. Despite the data being randomly sampled, there could be a possibility of voluntary bias since even after the GSS randomly selects its respondents, participation in the following survey is purely voluntary. 

Causality: Since this is an observational study, we cannot make any causal statements as no random assignment would be used. 

* * *

## Part 2: Research question

Question: Is there a significant difference in the religious beliefs of women aged 20-40 between the years 1980 and 2010?

Many societal norms and opinions have changed over the course of the last few decades in the United States, religion being one of them. As ways of thinking have evolved the world has become more progressive which has impacted religiuos beliefs and opinions. Since the driving force of any country is the age group of 20-40, a sample consisting of this group is being analyzed here. Also, a population of women is being chosen assumming that the societal norms with respect to women have had a more telling change than that of the male population from 1980 to 2010. 


* * *

## Part 3: Exploratory data analysis

First, we extract the population of interest as stated in the research question. In this case, the population of interest is women aged 20-40 years old in the years 1980 and 2010.

The code below extracts the population from the overall GSS dataset. 

```{r}
fem_sample <- gss %>%
  filter(sex=="Female") %>%
  filter(year == 1980 | year==2010) %>%
  filter(age>=20 & age<=40)
```

Since the variable of interest is "relig" which states what religion or none the individual follows, the distribution of this categorical variable can be plotted across the years and then across the ages for better understanding of the sample. 

Below is a plot of the various religious beliefs across the two years of interest - 1980 and 2010. 

```{r}
ggplot(data=fem_sample,aes(x=year,fill=relig))+geom_bar(binwidth = 10)
```

In order to see the proportion of religious views across the year, we need to create a mosaic plot with the same data as above. Since a mosaic plot consists of two categorical variables the "year" variable needs to be turned into a factor. 

Below, a new factor variable is added to the "fem_sample" dataframe which contains 1980 and 2010 as strings. 

```{r}
temp = factor(fem_sample$year,levels = c(1980,2010),labels = c("1980","2010"))

fem_sample <- fem_sample %>%
  mutate(year_cat = temp)

```

The mosaic plot below shows a proportional distribution of the religious views between the two years.

```{r}
plot(fem_sample$year_cat,fem_sample$relig)
```

Since the research question inquires about the individial having religious beliefs or not, the next step is to create a new variable which captures only whether a person has a specific religious following or None, as shown in the code below.

```{r}
temp = (fem_sample$relig=="None")
temp = factor(temp,levels=c("TRUE","FALSE"),labels = c("No","Yes"))
fem_sample <- fem_sample %>%
  mutate(relig_cat = temp)
```

Re-creating the barplot and the mosaic plot using the new categorical variable "religious", the code below shows us the new distributions. 

```{r}
ggplot(data=fem_sample,aes(x=year_cat,fill=relig_cat))+geom_bar()
```

```{r}
plot(fem_sample$year_cat,fem_sample$relig_cat,ylab="Religious",xlab="Year")
```

From the above plots, it appears that the proportion of women with non-religious views has increased from 1980 to 2010. 

Below are some summary statistics for the sample of data showing the number of religious and non-religious individuals grouped by year. 

Similar to the plots, the statistics below also point towards a greater rate of increase in women with non-religious views than women with religious views.

```{r}
fem_sample %>%
  filter(!(is.na(relig_cat))) %>%
  group_by(year,relig_cat) %>%
  summarise(n())
```

The proportion of women who appear to be non-religious in 1980 is 9.93% while that in 2010 is 25.50%.

The rate of increase for non-religious tendencies is approximately 169.69% whereas the increase for religious women is 5.12%. 

###Further Sampling

Taking 200 random samples from 1980 and 2010, and combining the observations for the inference section. Sampling equal number of observations for both categories helps in visualizing the proportions better while making the data more generalizable to the population. 

```{r}
fem_1980 <- fem_sample %>%
  filter(year==1980)

fem_2010 <- fem_sample %>%
  filter(year==2010)
```

```{r}
temp <- sample_n(fem_1980,200,replace=FALSE)
temp2 <- sample_n(fem_2010,200,replace=FALSE)
```

In the code above, 200 random samples are being extracted from the 1980 and 2010 categories.

Finally, a new dataset containing the randomly sampled observations are created below by combining the extracted samples. 

```{r}
fem_sample_200 <- rbind(temp,temp2)
```

The code below shows the new distribution of religious and non-religious female individuals across each year.

```{r}
fem_sample_200 %>%
  filter(!(is.na(relig_cat))) %>%
  group_by(year,relig_cat) %>%
  summarise(n())
```

The new proportion of non-religious individuals in 1980 is 9.29% while that in 2010 is 26.58%. Compared to the original dataset values of 9.93% and 25.50%, the proportions seem to have stayed about the same. 

* * *

## Part 4: Inference

###Confidence Interval

In this analysis, we are going to define a success as an individual not being religious and a failure as an individual being religious. 

Conditions 

1. Independence: For within groups (1980 and 2010), the observations are independent of each other due to random sampling being used for the data collection.  

We can safely assume that the sample size for each group is less than 10% of the entire population. 

Additionaly, for the between groups condition, the observations from 1980 and 2010 are likely different therefore we can assume these observations to be independent of each other. 

2. Sample size/skew: Since we are comparing proportions, we need to check the success-failure condition for each year - 1980 and 2010. 

Below, the sample sizes and success proportions for 1980 and 2010 respectively are defined.
```{r}
n1 = 200
n2 = 200
p1 = 17/n1 
p2 = 42/n2 
```

The results of the S-F condition for two proportions are listed below.
```{r}
n1*p1
n1*(1-p1)
n2*p2
n2*(1-p2)
```

All the results above satisfy the S-F condition for CI since they are all above 10. 

For this project, we will take a 95% confidence interval to determine if the CI contains the true population proportion of value of 0. 


```{r}
inference(y=relig_cat,x=as.factor(year),data=fem_sample_200,statistic="proportion",type="ci",method="theoretical",conf_level=0.95,success="No")
```

The sample proportion for 1980 is 0.085 while that for 2010 is 0.21. The accompanying plot shows the difference in proportion as well, similar to the plot shown in the EDA section. 

According to the rules of CI, if several samples of the same population of data were taken and this experiment repeated, 95% of the samples would contain the true population proportion. 

In this case, however, the CI of (-0.1934 , -0.0566) does not contain the proportion difference of 0. Therefore, we can consider the proportion difference of 0 to be an unusual occurrence. 

In fact, the CI generated above indicates that the proportion of women with non-religious tendencies is less in 1980 as compared to 2010 since the interval is completely on the negative side. 


###Hypotheses Testing

Null hypothesis: Since we are comparing two proportions and there is no previous assumption or belief regarding the difference in proportions, we can assume the null hypothesis to be that there has been no proportional change in the religious views of women aged 20-40 between 1980 and 2010. 

Alternate hypothesis: Religious views of women aged 20-40 between 1980 and 2010 has witnessed a proportional change. 

The conditions for conducting a HT for two proportions differs slightly from that for CI. The independence condition is the same but the S-F condition is determined differently.

For comparing the difference between two proportions, a pooled proportion is used to test the S-F condition. 

Pooled Proportion (p^): p^ = (total successes)/(total n)
```{r}
p_pooled = (17 + 42)/(200+200)
```
   
Success-Failure condition with pooled proportion:
```{r}
n1*p_pooled
n1*(1-p_pooled)
n2*p_pooled
n2*(1-p_pooled)
```

The S-F condition for a HT is met since all the above results are above 10. 

Since the conditions for HT are met, we continue to do the inference as shown below
```{r}
inference(y=relig_cat,x=as.factor(year),data=fem_sample_200,statistic="proportion",type="ht",method="theoretical",conf_level=0.95,success="No",null=0,alternative="twosided")
```

The above HT results show a very small p-value which is less than our significance level of 0.05. Therefore, we reject the null hypothesis. 

In other words, there is significant evidence that the proportion of non-religious women aged 20-40 has changed from 1980 to 2010. 

Interpreting in terms of probability, given that the proportion of women with non-religious views aged 20-40 is the same between 1980 and 2010, the probability of the observed or a more extreme result is 4e-04. This interpretation is well described in the second plot above containing the null distribution. 


###Conclusion

Both the CI as well as the HT inference testing provides significant evidence that the proportion of women aged 20-40 with non-religious views has changed from 1980 to 2010. 

On a closer look, the data indicates an increase in the proportion of non-religious views among the selected population of women in 2010. Hence we could say that there has been a statistically significant increase in non-religious views from 1980 to 2010.
